---
title: "Weather Events that Have Been of Greatest Impact to USA; Data from the U.S. 
National Oceanic and Atmospheric Administration's (NOAA)"
output:
  html_document:
    keep_md: true
---
  
  
##Introduction 

This is the second project assignment for the Reproducible Research Course.

##Synopsis

This project involves exploring the U.S. National Oceanic and Atmospheric 
Administration's (NOAA) storm database. We will be using primarily 5 variables:
  
- EVTYPE: event type
- FATALITIES: the number of fatalities corresponding to that event.
- INJURIES: the number of injuries people corresponding to that event.
- PROPDMG: the amount of money spent in properties damages.
- PROPDMGEXP: the prefix of the money spent (properties).
- CROPDMG: the amount of money spent in crops related damages.
- CROPDMGEXP: the prefix of the money spent (crops).

First we filtered rows that have as least one value greater than zero in one of 
the variables aforementioned. After that, used NOAA official event type categories
to kind of clean up the multiple typos in the variable EVTYPE in order to reduce
the amount of categories (it had over 900 unique values, when it supposes to have
48 categories.). Finally, we calculated the total fatalities, injuries people, 
total amount of money spent in properties and crops, and we proceed to graphs the
top 5 weather events with the most impact (either economic or population health)
to the USA.

##Data Processing

First of all, we load the packages required:

```{r packages, warning=FALSE}
#Load packages
pack <- c("ggplot2", "dplyr", "gridExtra")
usePackage <- function(p) {
        if (!is.element(p, installed.packages()[,1]))
                install.packages(p, dep = TRUE)
        require(p, character.only = TRUE)
}
sapply(pack, usePackage)
```

Then we load the dataset:

```{r data, cache=TRUE}
if(!file.exists(".\\data")) {dir.create(".\\data")}
fileurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!file.exists(".\\data\\NOAA_Data.csv.bz2")) {
        download.file(fileurl, destfile = ".\\data\\NOAA_Data.csv.bz2")
}

df <- read.csv(".\\data\\NOAA_Data.csv.bz2", stringsAsFactors = FALSE)
```

Now, let's take a brief look to the dataset:

```{r exploring}
head(df)
names(df)
str(df)
```
Now we subset rows with at least one value of the variables of interest greater
than 0: 

```{r subset}
df <- df[which(df$FATALITIES > 0 | df$INJURIES > 0 | df$CROPDMG > 0 | 
                       df$PROPDMG > 0), ]
```

After that, we create a dataset with the 48 categories according to 
[NOAA (page 6)][1]. The dataset with the 48 categories is available [here][2].
  
Let's load the data containing the categories:

```{r catData, cache=TRUE}
fileurl2 <- "https://raw.githubusercontent.com/kramus/RR_ProjectAssignment2/master/event.csv"
if(!file.exists(".\\data\\event.csv")) {
        download.file(fileurl2, destfile = ".\\data\\event.csv")
}
txt <- read.csv(".\\data\\event.csv", header = F, stringsAsFactors = FALSE)
#Change categories to upper cases
txt <- mutate_each(txt, funs(toupper))
```
 
Now in order to match the event type in the dataset with the official ones,
according to NOAA, we first need to change couple of values in the 
variable EVTYPE:

```{r changevalues}
df$EVTYPE[grep("TSTM WIND", df$EVTYPE)] <- "TROPICAL STORM"
df$EVTYPE[grep("^THUNDERSTORM", df$EVTYPE)] <- "THUNDERSTORM WIND"
df$EVTYPE[grep("SEVERE TH", df$EVTYPE)] <- "THUNDERSTORM"
df$EVTYPE[grep("HURRICANE", df$EVTYPE)] <- "HURRICANE/TYPHOON"
df$EVTYPE[grep("FLASH FL", df$EVTYPE)] <- "FLASH FLOOD"
df$EVTYPE[grep("FLOOD", df$EVTYPE)] <- "FLOOD"
df$EVTYPE[grep("HEAVY R", df$EVTYPE)] <- "HEAVY RAIN"
df$EVTYPE[grep("^HAIL", df$EVTYPE)] <- "HAIL"
df$EVTYPE[grep("HIGH W", df$EVTYPE)] <- "HIGH WIND"
df$EVTYPE[grep("TORNADO", df$EVTYPE)] <- "TORNADO"
df$EVTYPE[grep("TROPICAL ST", df$EVTYPE)] <- "TROPICAL STORM"
df$EVTYPE[grep("WILD F", df$EVTYPE)] <- "WILDFIRE"
df$EVTYPE[grep("WILDF", df$EVTYPE)] <- "WILDFIRE"
df$EVTYPE[grep("FIRE", df$EVTYPE)] <- "WILDFIRE"
df$EVTYPE[grep("EXTREME HEAT", df$EVTYPE)] <- "EXCESSIVE HEAT"
df$EVTYPE[grep("^HEAT", df$EVTYPE)] <- "HEAT"
df$EVTYPE[grep("WATERSP", df$EVTYPE)] <- "WATERSPOUT"
df$EVTYPE[grep("HEAVY SN", df$EVTYPE)] <- "HEAVY SNOW"
df$EVTYPE[grep("SNOW", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("COLD", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("Cold", df$EVTYPE)] <- "WINTER WEATHER"
df$EVTYPE[grep("WINTER ST", df$EVTYPE)] <- "WINTER STORM"
df$EVTYPE[grep("ICE", df$EVTYPE)] <- "ICE STORM"
df$EVTYPE[grep("^FREEZ", df$EVTYPE)] <- "FROST/FREEZE"
df$EVTYPE[grep("^FOG", df$EVTYPE)] <- "DENSE FOG"
```

Now we get the cleaner dataset as follows:

```{r cleandata}

cleandata <- df[df$EVTYPE %in% txt$V1, ]
```

In order to compare the percent of data excluded due typos or unknown event 
types: 

```{r typos}
df2 <- setdiff(df[, c(8, 23:28)], cleandata[, c(8, 23:28)])
#Percent of excluded rows
excl <- round(nrow(df2)/nrow(df)*100, digits = 2)
```

The percent of rows excluded due typos is ` r excl` which is great in my
opinion. 
  
Now that we have pass through data cleaning process, we can proceed to address 
the questions of interest.

##Results

####Question 1: Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

In order to assess the most harmful events with respect to population health, we need to calculate the sum of fatalities and injuries people among event types:

```{r q1}

q1 <- aggregate(cbind(FATALITIES, INJURIES) ~ 
                        EVTYPE, sum, data = cleandata)
topfatalities <- q1[order(q1$FATALITIES, decreasing = TRUE), c(1, 2)]
topfatalities <- topfatalities[1:5, ]
topinjuries <- q1[order(q1$INJURIES, decreasing = TRUE), c(1, 3)]
topinjuries <- topinjuries[1:5, ]
topfatalities
topinjuries
```

And the respective graph: 

```{r graph1}
#Graph 1
p1 <- ggplot(topfatalities, aes(x = reorder(EVTYPE, -FATALITIES), 
                                y = FATALITIES))
p1 <- p1 + geom_bar(stat = "identity", colour  = "red", fill = "chocolate2") +
        labs(x = "Event Type", y = "Number of Fatalities") +
        ggtitle(expression(atop("", 
                                atop("Fatalities", "")))) +
        theme(axis.title = element_text(face = "bold", size = 14), 
              axis.text.x = element_text(size = 8, angle = 30, hjust = 1, vjust = 1),
              plot.title = element_text(face = "bold", size = 20, vjust = 1.5), 
              legend.position = "none")
#Graph 2
p2 <- ggplot(topinjuries, aes(x = reorder(EVTYPE, -INJURIES), 
                                y = INJURIES))
p2 <- p2 + geom_bar(stat = "identity", colour  = "red", fill = "burlywood") +
        labs(x = "Event Type", y = "Number of Injuries") +
        ggtitle(expression(atop("", 
                atop("Injuries", "")))) +
        theme(axis.title = element_text(face = "bold", size = 14), 
              axis.text.x = element_text(size = 8, angle = 30, hjust = 1, vjust = 1),
              plot.title = element_text(face = "bold", size = 20, vjust = 1.5), 
              legend.position = "none")

#Join both graphs
grid.arrange(p1, p2, ncol=2, 
             top = "Top 5 Harmful Events Related to Population Health")
```

We can see that in both fatalities and injured people, the event with the most harmful with respect to population health is tornado.


####Question 2: Across the United States, which types of events have the greatest economic consequences?

In order to get the events that have the greatest economic consequences according
to NOAA dataset, we need to calculate the sum of the amount of money spent in 
properties and crops damages. First we create a multiplier prefix in order to get
the real total amount of money:

```{r economic}
multiprefix <- c("0" = 1, "K" = 1000, "k" = 1000, "M" = 1000000, "m" = 1000000, 
                 "H" = 100, "h" = 100, "5" = 100000, "6" = 1000000, "4" = 10000, 
                 "2" = 100, "7" = 10000000, "3" = 1000, "B" = 1000000000,
                 "b" = 1000000000)
```

There is an observation with a huge typo. It has a prefix of billions instead of
millions. The REFNUM is 605943. According to the [California Department of Water Resources][3], "In 2006, flood damage to the City of Napa and surrounding communities 
totaled approximately $115 million dollars; 1,200 homes and 250 businesses were 
damaged.". The dataset shows billions instead of millions. So we fix that error 
as follows:

```{r typo}
cleandata$PROPDMGEXP[which(cleandata$REFNUM == 605943)] <- "M"
```

Now let's create the multiplier prefix to calculate variable corresponding to 
the amount of money in damages:

```{r damages}

multiprefix <- c("0" = 1, "K" = 1000, "k" = 1000, "M" = 1000000, "m" = 1000000, 
                 "H" = 100, "h" = 100, "5" = 100000, "6" = 1000000, "4" = 10000, 
                 "2" = 100, "7" = 10000000, "3" = 1000, "B" = 1000000000,
                 "b" = 1000000000)

#Properties damages
cleandata$propdamage <- cleandata$PROPDMG * multiprefix[cleandata$PROPDMGEXP]

#Crops damages
cleandata$cropsdamage <- cleandata$CROPDMG * multiprefix[cleandata$CROPDMGEXP]

#Create data to answer question 2:
q2 <- aggregate(cbind(propdamage, cropsdamage) ~ 
                        EVTYPE, sum, data = cleandata)

#Getting top 5 event types with the greatest economic consequences
toppropdmg <- q2[order(q2$propdamage, decreasing = TRUE), c(1, 2)]
toppropdmg <- toppropdmg[1:5, ]
topcropsdmg <- q2[order(q2$cropsdamage, decreasing = TRUE), c(1, 3)]
topcropsdmg <- topcropsdmg[1:5, ]
toppropdmg
topcropsdmg
```

Finally, let's construct the respective graphs:

```{r graphs2}

#Graph 1
p11 <- ggplot(toppropdmg, aes(x = reorder(EVTYPE, -propdamage), 
                                y = propdamage))
p11 <- p11 + geom_bar(stat = "identity", colour  = "red", fill = "chocolate2") +
        labs(x = "Event Type", y = "Amount of Money Spent") +
        ggtitle(expression(atop("", 
                                atop("Properties Damages", "")))) +
        theme(axis.title = element_text(face = "bold", size = 14), 
              axis.text.x = element_text(size = 8, angle = 30, hjust = 1, 
                                         vjust = 1),
              plot.title = element_text(face = "bold", size = 20, vjust = 1.5), 
              legend.position = "none")

#Graph 2
p22 <- ggplot(topcropsdmg, aes(x = reorder(EVTYPE, -cropsdamage), 
                              y = cropsdamage))
p22 <- p22 + geom_bar(stat = "identity", colour  = "red", fill = "burlywood") +
        labs(x = "Event Type", y = "Amount of Money Spent") +
        ggtitle(expression(atop("", 
                                atop("Crops Damages", "")))) +
        theme(axis.title = element_text(face = "bold", size = 14), 
              axis.text.x = element_text(size = 8, angle = 30, hjust = 1, 
                                         vjust = 1),
              plot.title = element_text(face = "bold", size = 20, vjust = 1.5), 
              legend.position = "none")

#Join both graphs
grid.arrange(p11, p22, ncol=2, 
             top = "Top 5 Expensive Natural Events in the USA")
```

We can see that for properties and crops damages, the event with the greatest
economic consequences is hurricane/typhoon and flood respectively. 

[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf "NOAA (page 6)"
[2]: https://raw.githubusercontent.com/kramus/RR_ProjectAssignment2/master/event.csv "here"
[3]: http://www.water.ca.gov/floodsafe/ca-flood-preparedness/fpw-day3.cfm "California Department of Water Resources"
